### 2022-05-27 paper/localpsf.tex
wrote computational cost section of paper. Ready for editing pass

The computational cost of our approximation method may be divided into two categories: (1) the cost of building the H-matrix approximation, and (2) the cost of using the H-matrix approximation to perform further linear algebra operations. The cost to build the H-matrix approximation may further be broken down into (1a) the cost of computing the impulse response moments and batches, which involves applying $\Aop$ to vectors, and (1b) the cost of building the H-matrix after the impulse response moments and batches have been computed, which does not involve applying $\Aop$ to vectors. In practice, (1a) is the dominant cost, because our method is targeted at applications in which applying the operator $\Aop$ to a vector involves performing an expensive computational procedure such as solving a large-scale partial differential equation. Conceptually, our method was designed to construct an approximation that is as accurate as possible, using as few operator applies as possible, subject to the constraint that the further costs associated with building and using the approximation are near-linear (and therefore scalable) in $N$.

\paragraph{(1a): Impulse response moments and batches} Computing $V$, $\mu$, and $\Sigma$ require applying $\Aop$ to $1$, $d$, and $d(d+1)/2$ vectors, respectively. Concretely, in one spatial dimensions ($d=1$), computing $V$, $\mu$, and $\Sigma$ requires $1+1+1=3$ operator applies. In two spatial dimensions, this cost is $1+2+3=6$ operator applies. In three spatial dimensions, this cost is $1+3+6=10$ operator applies. Computing each impulse response batch requires applying $\Aop$ to $1$ vector, so computing the impulse response batches $\{\eta_i\}_{i=1}^{n_b}$ requires $n_b$ operator applies. In total, computing the impulse response moments and batches requires
\begin{equation*}
	1 + d + d(d+1)/2 + n_b \qquad \text{operator applies.}
\end{equation*}
In a typical application one might have $d=2$ and $n_b=5$, in which case $11$ operator applies would be required. Choosing one sample point batch via the greedy ellipsoid packing algorithm described in Section \ref{sec:sample_point_selection} may require performing up to $Ns$ ellipsoid intersection tests, where $s$ is the number of sample points in the batch. Choosing all of the sample points therefore requires performing at most
\begin{equation*}
	Nm \qquad \text{ellipsoid intersection tests,}
\end{equation*}
where $m$ is the total number of sample points in all batches. The multiplicative dependence of $N$ with $m$ is undesirable, and reducing this cost is a target for future work. The cost of picking sample points could be reduced, for example, by ``thinning'' the sample points. That is, choosing batches from a smaller subset of $N_\text{thin} << N$ points, rather than all $N$ finite element Lagrange nodes. The cost of picking one batch could be reduced from $N s$ to $O(N \log s)$ using more advanced computational geometry techniques, such as organizing the ellipsoids associated with points that have already been picked in a batch into an AABB tree, then using this AABB tree to accelerate the required ellipsoid intersection checks. This would reduce the cost of choosing all sample point batches from $Nm$ to $O(N n_b \log s)$. However, in practice the cost for picking sample points is small compared to other parts of the approximation algorithm, so we do not pursue these improvements in this paper.

\paragraph{(1b): Building the H-matrix} The classical H-matrix construction technique, which is described in Appendix \ref{app:h_matrix}, requires evaluating $O(k_h^2 N \log N)$ matrix entries of the approximation, where $k_h$ is the H-matrix rank (maximum rank among the blocks of the H-matrix). To evaluate one matrix entry, first one must perform $k_n$ nearest neighbors queries of the sample points KD tree, and each of these queries requires $O(\log m)$ work. Next, one must finding the mesh cells that $k_n$ distinct points reside in using our mesh AABB tree, which requires $O(k_n \log N)$ work, and evaluate finite element basis functions on each of those cells, which requires $O(k_n)$ work. Finally, one must solve a $k_n \times k_n$ linear system to perform the radial basis function interpolation, which requires $O(k_n^3)$ work. Overall, building the H-matrix therefore requires
\begin{equation*}
	O\left(\left(k_h^2 N \log N\right)\left(k_n \log N + k_n^3\right)\right) \qquad \text{work}.
\end{equation*}
This cost is nearly linear in $N$, scalable with respect to the number of sample points $m$, and quite fast in practice.

In our view, the goal of the approximation method is to reduce the number of operator applies as much as possible

we consider the operator applies to be the dominant cost to a vector is an expensive computational procedure. For example, in our numerical results we demonstrate the method for applications in which applying $\Aop$ to a vector requires solving two partial differential equations, which are defined on a domain with a larger spacetime dimension than that of $\Omega$.

WORK IN PROGRESS BELOW HERE!!!

\begin{tabular}{lll}
	Variable name & Approximate size & Variable \\
	\hline
	$N$ & $10^3$--$10^9$ & Number of finite element degrees of freedom \\
	$b$ & $5$--$25$ & number of impulse response batches \\
	$r$ & $10$--$30$ & H-matrix rank \\
	$k$ & $5$--$15$ & number of nearest neighbors \\
	$\{\eta_b\}_{b=1}^\text{num\_batches}$ & $\text{num\_batches}$ actions of $\mathcal{A}$
\end{tabular}

$k_\text{hmatrix}$, $k_\text{nbrs}$, $k_\text{batches}$

\begin{tabular}{ll}
	$V$ & $1$ action of $\mathcal{A}^T$ \\
	$\mu$ & $d$ actions of $\mathcal{A}^T$ \\
	$\Sigma$ & $d(d+1)/2$ actions of $\mathcal{A}^T$ \\
	$\{\eta_b\}_{b=1}^\text{num\_batches}$ & $\text{num\_batches}$ actions of $\mathcal{A}$
\end{tabular}

Computing $V$, $\mu$, and $\Sigma$ require $1$, $d$, and $d(d+1)/2$ operator actions of $\Aop$, respectively. Computing the impulse response batches $\eta_b$ requires one operator action of $\Aop$ per batch. Picking the sample points

In the non-symmetric case, evaluating $\widetilde{\Phi}(y,x)$ requires us to perform one $k$-nearest neighbor query of the KD-tree of sample points, $k$ point collision queries of the AABB-tree of mesh cells, and solve a $k \times k$ linear system. In the symmetric case, we must perform two $k$-nearest neighbor queries of the KD-tree of sample points, $2k$ point collision queries of the AABB-tree of mesh cells, and solve a $4k \times 2k$ linear least squares problem via QR factorization. Both of these cases require $O(k \log r + k \log N_\text{cell} + k^3)$ work, where $r$ is the number of sample points, and $N_\text{cell}$ is the number of cells in the mesh used to discretize the problem. Since $k$ is small, this is cheap.

>>>>

The computational cost of our approximation method may be divided into two categories: (1) the cost of building the H-matrix approximation, and (2) the cost of using the H-matrix approximation to perform further linear algebra operations. The cost to build the H-matrix approximation may further be broken down into (1a) the cost of computing the impulse response moments and batches, which involves applying $\Aop$ to vectors, and (1b) the cost of building the H-matrix after the impulse response moments and batches have been computed, which does not involve applying $\Aop$ to vectors. In practice, (1a) is the dominant cost, because our method is targeted at applications in which applying the operator $\Aop$ to a vector involves performing an expensive computational procedure such as solving a large-scale partial differential equation. Conceptually, our method was designed to construct an approximation that is as accurate as possible, using as few operator applies as possible, subject to the constraint that the further costs associated with building and using the approximation are near-linear (and therefore scalable) in $N$. We now describe these costs in detail. For convenience, variable symbols and their approximate sizes are listed in Table \ref{tab:vars}.

\begin{table}
	\begin{tabular}{lll}
		Symbol & Approximate size & Variable name \\
		\hline
		$N$ & $10^2$--$10^9$ & Number of finite element degrees of freedom \\
		$n_b$ & $1$--$25$ & Number of impulse response batches \\
		$k_h$ & $5$--$50$ & H-matrix rank \\
		$k_n$ & $5$--$15$ & Number of nearest neighbors for RBF interpolation \\
		$d$ & $1$--$3$ & Spatial dimension \\
		$r$ & $10^1$--$10^4$ & Classical matrix rank \\
		$m$ & $10^1$--$10^4$ & Total number of sample points (all batches) \\
		$s$ & $1$--$500$ & Number of sample points in one batch \\
		$l$ & $1$--$2$ & Rational function parameter for SPD modification
	\end{tabular}
	\caption{Symbols used for variables in computational cost estimates, and their approximate sizes.}
	\label{tab:vars}
\end{table}

\paragraph{(1a): Impulse response moments and batches} Computing $V$, $\mu$, and $\Sigma$ require applying $\Aop$ to $1$, $d$, and $d(d+1)/2$ vectors, respectively. Concretely, in one spatial dimensions ($d=1$), computing $V$, $\mu$, and $\Sigma$ requires $1+1+1=3$ operator applies. In two spatial dimensions, this cost is $1+2+3=6$ operator applies. In three spatial dimensions, this cost is $1+3+6=10$ operator applies. Computing each impulse response batch requires applying $\Aop$ to $1$ vector, so computing the impulse response batches $\{\eta_i\}_{i=1}^{n_b}$ requires $n_b$ operator applies. In total, computing the impulse response moments and batches therefore requires
\begin{equation*}
	1 + d + d(d+1)/2 + n_b \qquad \text{operator applies.}
\end{equation*}
In a typical application one might have $d=2$ and $n_b=5$, in which case $11$ operator applies would be required. In contrast, matrix-free low-rank approximation methods require $O(r)$ operator applies, where $r$ is the classical numerical rank of $\Aop$. Since $r$ may be in the thousands, low rank approximation may require thousands of operator applies.

Computing the impulse response batches also requires choosing sample point batches via the greedy ellipsoid packing algorithm described in Section \ref{sec:sample_point_selection}. Choosing one batch of sample points may require performing up to $Ns$ ellipsoid intersection tests, where $s$ is the number of sample points in the batch. Choosing all of the sample points therefore requires performing at most
\begin{equation*}
	Nm \qquad \text{ellipsoid intersection tests,}
\end{equation*}
where $m$ is the total number of sample points in all batches. The multiplicative dependence of $N$ with $m$ is undesirable, and reducing this cost is a target for future work. The cost of picking sample points could be reduced, for example, by ``thinning'' the set of candidate points that the sample points are picked from. That is, choosing sample points from a smaller subset of $N_\text{thin} << N$ points, rather than all $N$ finite element Lagrange nodes. The cost of picking one batch could be reduced from $N s$ to $O(N \log s)$ using more advanced computational geometry techniques, such as organizing the ellipsoids associated with points that have already been picked in a batch into an AABB tree, then using this AABB tree to accelerate the required ellipsoid intersection checks. This would reduce the cost of choosing all sample point batches from $Nm$ to $O(N n_b \log s)$. In practice the cost for picking sample points is small compared to other parts of the approximation algorithm, so we do not pursue these improvements in this paper.

\paragraph{(1b): Building the H-matrix} The classical H-matrix construction technique, which is described in Appendix \ref{app:h_matrix}, requires evaluating $O(k_h^2 N \log N)$ matrix entries of the approximation, where $k_h$ is the H-matrix rank (maximum rank among the blocks of the H-matrix). To evaluate one matrix entry, first one must find the $k_n$ nearest sample points to a given point, where $k_n$ is the number of impulses used in the RBF interpolation. This is done using a precomputed KD tree of sample points, and requires $O(k_n \log m)$ work. Next, one must find the mesh cells that the points $\{z_i\}_{i=1}^{k+n}$ reside. This is done using our mesh AABB tree, and requires $O(k_n \log N)$ work. Next, we must evaluate finite element basis functions on each of those cells, which requires $O(k_n)$ work. Finally, we must solve a $k_n \times k_n$ linear system to perform the radial basis function interpolation, which requires $O(k_n^3)$ work. Overall, building the H-matrix therefore requires
\begin{equation*}
	O\left(\left(k_h^2 N \log N\right)\left(k_n \log N + k_n^3\right)\right) \qquad \text{work}.
\end{equation*}
This cost is nearly linear in $N$, scalable with respect to the number of sample points $m$, and quite fast in practice.

\paragraph{(2): H-matrix operations} H-matrix methods for matrix-vector products, matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion require
\begin{equation*}
	O(k_h^2 N \log(N)^a) \qquad \text{work},
\end{equation*}
where $a \in \{0,1,2,3\}$. The precise cost depends on the type of H-matrix used and the operation being performed; for more details see \cite{GrasedyckHackbusch03}. The rational method for modifying $\mathbf{A}_H$ to be positive semi-definite, described in Appendix \ref{sec:make_spd}, is performed using two matrix-matrix additions, $l+1$ matrix-matrix multiplications, and one matrix inversion, where $2^l$ is the order of the rational function used.

%$k_\text{hmatrix}$, $k_\text{nbrs}$, $k_\text{batches}$
%
%\begin{tabular}{ll}
%	$V$ & $1$ action of $\mathcal{A}^T$ \\
%	$\mu$ & $d$ actions of $\mathcal{A}^T$ \\
%	$\Sigma$ & $d(d+1)/2$ actions of $\mathcal{A}^T$ \\
%	$\{\eta_b\}_{b=1}^\text{num\_batches}$ & $\text{num\_batches}$ actions of $\mathcal{A}$
%\end{tabular}
%
%
%Computing $V$, $\mu$, and $\Sigma$ require $1$, $d$, and $d(d+1)/2$ operator actions of $\Aop$, respectively. Computing the impulse response batches $\eta_b$ requires one operator action of $\Aop$ per batch. Picking the sample points 
%
%In the non-symmetric case, evaluating $\widetilde{\Phi}(y,x)$ requires us to perform one $k$-nearest neighbor query of the KD-tree of sample points, $k$ point collision queries of the AABB-tree of mesh cells, and solve a $k \times k$ linear system. In the symmetric case, we must perform two $k$-nearest neighbor queries of the KD-tree of sample points, $2k$ point collision queries of the AABB-tree of mesh cells, and solve a $4k \times 2k$ linear least squares problem via QR factorization. Both of these cases require $O(k \log r + k \log N_\text{cell} + k^3)$ work, where $r$ is the number of sample points, and $N_\text{cell}$ is the number of cells in the mesh used to discretize the problem. Since $k$ is small, this is cheap.
